Concrete Semantics for Pushdown Analysis:\\The Essence of Summarization

### 2013-10-02 content.tex
Moved proofs to appendix, reformatted sentences to be on single lines for better diffs, added GC section.

Programs in higher-order languages heavily use function calls and
method dispatch as part of their control flow. Until recently, flow
analyses for higher-order languages could not handle return flow
precisely \citep{ianjohnson:vardoulakis-lmcs11,
  ianjohnson:earl2010pdcfa}, which leads to several spurious paths
(and thus false positives) due to the pervasiveness of function/method
calls and subsequent returns. These works, called CFA2 and PDCFA
respectively, use pushdown automata as their approximation's target
model of computation. They are hence called ``pushdown
analyses.''\footnote{We will refer to the classic finite model
  analyses as ``regular analyses'' after regular languages.} CFA2 and
PDCFA have difficult details to easily apply to an off-the-shelf
semantics --- especially if they feature non-local control transfer
that breaks the pushdown model.

There is a systematic process for transforming off-the-shelf
programming language semantics into a form amenable to \emph{regular}
analysis that has been widely applied with great success to production
programming languages ~\citet{dvanhorn:VanHorn2010Abstracting} (a
technique called abstracting abstract machines, or AAM). A
contribution of this paper is a systematic process to construct
\emph{pushdown} analyses of programming languages, due to the
precision (and often performance) benefits.

Testing new ideas in analysis for improving precision or increasing
performance can be a difficult venture. We contend that the machinery
that we employ in the abstract should have a concrete counterpart that
maintains the meaning of the language so that we can see their effect
without introducing approximations which might mask correctness
issues. Abstraction should be a simple process that is ``obviously
correct.'' The framework we present in this paper is applicable in the
concrete such that a point-wise abstraction leads to the pushdown
analyses in the literature.

This paper gives a common, simple framework to derive both CFA2 and
PDCFA using a recipe to apply pushdown analysis techniques to
arbitrary semantics in a manner similar to AAM. To exercise the recipe further, 
we give a novel analysis for delimited, composable control. \\

@@

CFA2 uses a technique called ``summarization'' from
\citet[Chapter 7]{local:muchnick:jones:flow-analysis:1981}, which is synonymous with the
$\epsilon$-closure graph that PDCFA constructs. Summarization
algorithms need not be restricted to languages with well-bracketed
calls and returns. We can adopt the technique for higher precision in
the common case but still handle difficult cases such as first-class
control. This was shown for the
\rackett{call-with-current-continuation} (a.k.a. \rackett{call/cc})
operator in ~\citet{ianjohnson:Vardoulakis2011Pushdown}. This
impressive work illuminated the fact that we can harness the enhanced
technology of pushdown analyses in non-pushdown models of
computation. Doing this sacrifices call/return matching in the general
case, but in practice the precision is much better than the
alternative regular model that, say, AAM would provide.

A downside of the work providing \rackett{call/cc} is that it was an
algorithmic change to the already complicated CFA2 --- there was no
recipe for how to do this for one's favorite control operator. This
paper seeks to do just that with an operational view of what
summarization is, in essence. In other words, we give a concrete
semantics to the tricks that the analyses in the literature use in the
abstract, and maintain the original meaning of the language. We also
give intuitive analogies to well-established ideas/techniques so that
the working semanticist can write a pushdown analysis for her
language. In order to demonstrate the applicability of this viewpoint,
we show a new analysis for a language with composable control. All of
the semantics modeled in this paper are implemented in full detail in
PLT Redex~\citep{ianjohnson:Felleisen:2009:SEP:1795772} and available
online\footnote{\url{http://github.com/ianj/concrete-summaries}}.

@@

PDCFA does not have some orthogonal semantic components that CFA2
features to improve precision and is thus the simpler of the two. The
key to their method is to notice in languages without stack-capturing
features, the continuation is only modified a frame at a time. If the
state space without the stack is finite (and it can be made finite
with an approximation ala AAM), then the model falls directly into the
realm of pushdown systems. They thus recast the problem in terms of
pushdown systems. The whole of their machinery is then computing the
pushdown system on-the-fly, only considering states reachable from the
root (initial) state. They call the main data-structure for this a
Dyck state graph. To match nodes that push frames with later nodes
following the pop of that frame, they additionally compute an
$\epsilon$-closure graph, which keeps edges between nodes that have a
net-zero stack change path between them. We skip the detour to
pushdown systems altogether and show an operational understanding of
the resulting analysis.

We start by deriving PDCFA from an operational semantics for the
untyped lambda calculus (figure \ref{fig:base-semantics}). The
semantic spaces for the machine follow.

\begin{align*}
  \mexpr \in \Expr &= \svar{\mvar} \mid \sapp{\mexpr}{\mexpr} \mid \slam{\mvar}{\mexpr} \\
  \mstate \in \State &= (\Expr \times \Env) \times \Store \times \Kont \\
  \mval \in \Value &::= \vclo{\slam{\mvar}{\mexpr}}{\menv} \\
  \mkont \in \Kont &::= \kmt \mid \kar{\mexpr,\menv,\mkont} \mid \kfn{\mval,\mkont} \\
  \menv \in \Env &= \Var \to \Addr \\
  \mstore \in \Store &= \Addr \to \wp(\Value) \\
  \mvar \in \Var &\text{ an infinite set} \\
  \maddr \in \Addr &\text{ an infinite set}
\end{align*}

The $\Addr$ space is what controls the precision of the model. For a
concrete semantics, we require the allocation meta-function $\alloc :
\State \to \Addr$ to return fresh addresses ($\maddr \notin
\dom(\mstore)$), but any choice of address is sound. If $\alloc$ only
uses addresses from a finite subset of $\Addr$, then the state space
without continuations is finite, and the space of contination frames
is finite (thus the pushdown system interpretation is apt). This
meta-function along with the commitment to having no recursive
data-structures is the key to the AAM technique. All recursion can be
expressed by indirecting through addresses into a store --- tying
Landin's knot. We borrow this technique on top of our own.

\begin{figure}
  \centering
  $\mstate \stepto \mstate' \text{ where } \maddr = \alloc(\mstate)$ \\
  \begin{tabular}{r|l}
    \hline
% variable lookup
    $\tpl{(\svar\mvar, \menv), \mstore, \mkont}$
    &
    $\tpl{\mval,\mstore,\mkont}$ if $\mval \in \mstore(\menv(\mvar))$
    \\
% application
    $\tpl{(\sapp{\mexpri0}{\mexpri1}, \menv), \mstore, \mkont}$
    &
    $\tpl{(\mexpri0, \menv), \mstore, \kar{\mexpri1,\menv,\mkont}}$
    \\
% argument evaluation
    $\tpl{\mval, \mstore, \kar{\mexpr,\menv,\mkont}}$
    &
    $\tpl{(\mexpr, \menv), \mstore, \kfn{\mval, \mkont}}$
    \\
% function call
    $\tpl{\mval,\mstore,\kfn{\vclo{\slam{\mvar}{\mexpr}}{\menv}, \mkont}}$
    &
    $\tpl{(\mexpr, \extm{\menv}{\mvar}{\maddr}), \joinone{\mstore}{\maddr}{\mval}, \mkont}$
  \end{tabular}
  \caption{The CESK machine}
  \label{fig:base-semantics}
\end{figure}

\paragraph{Tracking return points:} the magic of the method is in
keeping a table of continuation segments for each function and store
pair, where the continuation is segmented at function boundaries. This
closely mirrors the technique of store-allocating continuations used
in AAM. Instead of deferring to the table (store) for the remainder of
a continuation for each frame, we only have indirections at function
call boundaries (see rules 4 and 5 in figure
\ref{fig:summary-semantics} for saving and restoring
continuations). Note that since AAM is about finitizing the state
space, this step itself fits within the AAM method, since
continuations within functions truncated at call boundaries are
bounded by the nesting depth of those functions, thereby making the
continuation space finite. In the monovariant case, the number of
continuations is still linear in the size of the program.

The tail of a continuation in the context of a function will contain
the stack-less context in which the function was called, in order to
link up with the proper call-site(s). The context includes the
function (or unique label of the function), the environment, and the
store. Notice that this choice can be viewed as a special allocation
strategy for continuations in the AAM viewpoint, but we separate the
table of continuations from the store since continuations themselves
contain stores in $\rt$ frames --- this leads to a recursive
data-structure that we are trying to avoid. The store is an important
ingredient to maintaining enough precision to keep a pushdown
abstraction. The semantic spaces for the machine are modified thusly:

\begin{align*}
  \mstate \in \State &= (\Expr \times \Env) \times \Store \times \Kont \times \KTab \times \Memo \\
  \mkont \in \Kont &::= \kmt
                     \mid \kar{\mexpr,\menv,\mkont} 
                     \mid \kfn{\mval,\mkont}
                     \mid \krt{(\mexpr,\menv),\mstore}\\
  \mktab \in \KTab &= (\Expr \times \Env)\times \Store \to \wp(\Kont) \\
  \mmemo \in \Memo &= (\Expr \times \Env)\times \Store \to \wp(\Value \times \Store)
\end{align*}

\paragraph{The role of summaries:} notice the additional $\Memo$
component. A ``summary edge'' in CFA2 or equivalently an
$\epsilon$-edge in PDCFA is an edge from the source of a push edge to
the target of the matching pop edge. ``Matching'' here means there is
a path through machine reductions that don't change the stack, or
through summary edges. In our model, we only ``push'' when we call a
function, and we only ``pop'' when we return with a value. There is an
analogy to something more operational: summary edges embody
memoization (see the last rule in figure
\ref{fig:summary-semantics}). Instead of following an entire path
through a call to return a value, we simply jump from the call to the
return with the result of the call, with the resulting context changes---in this case only the store (the second case of rule 4).

\begin{figure}
  \centering
  $\mstate \stepto \mstate' \text{ where } \maddr = \alloc(\mstate)$ \\
  \begin{tabular}{r|l}
    \hline
% variable lookup
    $\tpl{(\svar\mvar, \menv), \mstore, \mkont, \mktab, \mmemo}$
    &
    $\tpl{\mval,\mstore,\mkont, \mktab, \mmemo}$ if $\mval \in \mstore(\menv(\mvar))$
    \\
% application
    $\tpl{(\sapp{\mexpri0}{\mexpri1}, \menv), \mstore, \mkont, \mktab, \mmemo}$
    &
    $\tpl{(\mexpri0, \menv), \mstore, \kar{\mexpri1,\menv,\mkont}, \mktab, \mmemo}$
    \\
% argument evaluation
    $\tpl{\mval, \mstore, \kar{\mexpr,\menv,\mkont}, \mktab, \mmemo}$
    &
    $\tpl{(\mexpr, \menv), \mstore, \kfn{\mval, \mkont}, \mktab, \mmemo}$
    \\
% function call
    $\tpl{\mval,\mstore,\kfn{\vclo{\slam{\mvar}{\mexpr}}{\menv}, \mkont},\mktab,\mmemo}$
    & % actually do call,
    $\tpl{\mpoint,
          \mstore',
          \krt{\mpoint, \mstore'},
          \mktab',
          \mmemo}$
\\
    & or \\
    & $\tpl{\mval_\mathit{result},
            \mstore'',
            \mkont,
            \mktab',
            \mmemo}$
    \\ & \quad if $(\mval_\mathit{result},\mstore'') \in \mmemo(\mpoint,\mstore')$
    \\ % or, lookup in table
%    & $\tpl{\mval, \mstore', \mkont,\mktab,\mmemo}$ if $\mval \in \mmemo(\mpoint, \mstore')$ \\    
    where & $\mpoint = (\mexpr, \extm{\menv}{\mvar}{\maddr})$ \\
          & $\mstore' = \joinone{\mstore}{\maddr}{\mval}$ \\
          & $\mktab' = \joinone{\mktab}{(\mpoint, \mstore')}{\mkont}$
    \\
% return
    $\tpl{\mval, \mstore, \krt{\mpoint,\mstore'}, \mktab, \mmemo}$
    &
    $\tpl{\mval, \mstore, \mkont, \mktab, \joinone{\mmemo}{(\mpoint, \mstore')}{(\mval,\mstore)}}$
    \\ & \quad if $\mkont \in \mktab(\mpoint, \mstore')$
  \end{tabular}
  \caption{The summarizing tabular stack machine}
  \label{fig:summary-semantics}
\end{figure}

To turn this semantics into PDCFA's algorithm for constructing a Dyck
state graph, given the appropriate $\alloc$, we apply a widening
operator to make $\mstore$, $\mkont$, and $\mmemo$ shared amongst all
states (meaning states then become represented without these
compontents), in what is then called a $\System$.  Since these
components are shared, they are always the least upper bound of any
respective component that the semantics produces (the intermediate set
$I$) in order to stay sound. We use a meta-function $\wn$, ``wide to
narrow'', to shift between the two representations of states.

%% FIXME: remove stores from memo-table since they will always be unchanged
%% TODO: Show that store widening and store/memo/KTab widening have a bisimulation.
\begin{align*}
  \mastate \in \sa{State} &= (\Expr \times \Env) \times \Kont \\
  \System &= \wp(\sa{State} \times \Store) \times \wp(\sa{State}) \\
          &\phantom{=} \times \Store \times \KTab \times \Memo \\
  {\mathcal F} &: \System \to \System \\
  {\mathcal F}(S, F, \mstore, \mktab, \mmemo) &= (S \cup S', F', \mstore', \mktab', \mmemo') \\
  \text{where } I &= \setbuild{\mstate'}{\mastate \in F, \wn(\mastate,\mstore,\mktab,\mmemo) \stepto \mstate'} \\
                \mstore' &= \bigsqcup\setbuild{\mstore'}{\wn(\_, \mstore',\_,\_) \in I} \\
                \mktab' &=  \bigsqcup\setbuild{\mktab'}{\wn(\_, \_, \mktab', \_) \in I} \\
                \mmemo' &=  \bigsqcup\setbuild{\mmemo'}{\wn(\_, \_, \_, \mmemo') \in I} \\
                S' &= \setbuild{(\mastate',\mstore')}{\wn(\mastate', \_, \_, \_) \in I} \\
                F' &= \setbuild{\mastate'}{(\mastate',\_) \in S' \setminus S} \\
                \wn(\tpl{(\mexpr,\menv),\mkont},\mstore,\mktab,\mmemo)
                   &= \tpl{(\mexpr,\menv),\mstore,\mkont,\mktab,\mmemo}
\end{align*}

A $\System$ embodies the states seen and at which store, $S$, the
frontier set of states yet to analyze, $F$, the shared store for the
frontier, $\mstore$, the continuation table $\mktab$ and the memo
table $\mmemo$. The memo table need only map to values, since the
global store will always subsume the store that would otherwise be
mapped. All the states in the frontier are stepped with the current
store, after which the next store is the least upper bound of all the
resulting stores. The next frontier contains only states resulting
from stepping the previous frontier that we haven't seen at this next
store. Systematic techniques for a performant implementation can be
found in \citet{ianjohnson:oaam:icfp2013}.

\paragraph{Example:} Let's consider a monovariant allocation strategy
(the address for each binding is the variable name itself) to run the
following example:
\begin{lstlisting}[mathescape]
  (let* ([id ($\lambda$ (x) x)]
         [y (id 0)]
         [z (id 1)])
    ($\le$ y z))
\end{lstlisting}
Suppose we extend our semantics to allow numbers, numeric primitives
and \texttt{let}. The continuation frame for a \texttt{let} contains
the identifier to bind to the resulting value, along with the body of
the \texttt{let} with its environment. Call the constructor of this
frame $\mathbf{lt}$.

We should expect that a pushdown analysis would predict this evaluates
to true, and there are no loops in the program. 0CFA ~\citep{ianjohnson:Shivers:1991:CFA} claims there is a
loop from the second call of \texttt{id} to the first, and thusly
predicts this program evaluates to true or false. PDCFA claims there
is no loop, and depending on the implementation, that the result is
true or false (paper), or just true (implemented)\footnote{This is
  because the paper steps every seen state with the current store
  every iteration, but the implementation only steps states that need stepping.}. Let's take a look at the evaluation after binding
\texttt{id} ($\mstore_0 = [\texttt{id} \mapsto
\set{\vclo{\slam{\texttt{x}}{\texttt{x}}}{\bot}}]$):

@@

We maintained enough context to distinguish the return points of
\texttt{id} to not rebind \texttt{y} to 1. When determining control
flow through the expression, we consult $\mmemo$ to continue past
function calls, so there is no confusion about a back edge from the
second call to \texttt{id}.

\begin{conjecture}
  Tabular semantics has a bisimulation with the standard semantics with a sound allocation strategy.
\end{conjecture}
\begin{proof}

\end{proof}

\begin{conjecture}
  Store-widened tabular semantics has a bisimulation with store/memo/k-table-widened semantics.
\end{conjecture}

>>>>

Programs in higher-order languages heavily use function calls and method dispatch as part of their control flow.
%
Until recently, flow analyses for higher-order languages could not handle return flow precisely \citep{ianjohnson:vardoulakis-lmcs11, ianjohnson:earl2010pdcfa}, which leads to several spurious paths (and thus false positives) due to the pervasiveness of function/method calls and subsequent returns.
%
These works, called CFA2 and PDCFA respectively, use pushdown automata as their approximation's target model of computation.
%
They are hence called ``pushdown analyses.''
%
\footnote{We will refer to the classic finite model analyses as ``regular analyses'' after regular languages.}
%
CFA2 and PDCFA have difficult details to easily apply to an off-the-shelf semantics --- especially if they feature non-local control transfer that breaks the pushdown model.
%%

%%
There is a systematic process for transforming off-the-shelf programming language semantics into a form amenable to \emph{regular} analysis that has been widely applied with great success to production programming languages ~\citet{dvanhorn:VanHorn2010Abstracting} (a technique called abstracting abstract machines, or AAM).
%
A contribution of this paper is a systematic process to construct \emph{pushdown} analyses of programming languages, due to the precision (and often performance) benefits.
%%

%%
Testing new ideas in analysis for improving precision or increasing performance can be a difficult venture.
%
We contend that the machinery that we employ in the abstract should have a concrete counterpart that maintains the meaning of the language so that we can see their effect without introducing approximations which might mask correctness issues.
%
Abstraction should be a simple process that is ``obviously correct.''
%
The framework we present in this paper is applicable in the concrete such that a point-wise abstraction leads to the pushdown analyses in the literature.
%%

%%
This paper gives a common, simple framework to derive both CFA2 and PDCFA using a recipe to apply pushdown analysis techniques to arbitrary semantics in a manner similar to AAM.
%
To exercise the recipe further, we give a novel analysis for delimited, composable control. \\

@@

CFA2 uses a technique called ``summarization'' from Sharir and Pnueli \citep[Chapter 7]{local:muchnick:jones:flow-analysis:1981}, which is synonymous with the $\epsilon$-closure graph that PDCFA constructs.
%
Summarization algorithms need not be restricted to languages with well-bracketed calls and returns.
%
We can adopt the technique for higher precision in the common case but still handle difficult cases such as first-class control.
%
This was shown for the \rackett{call-with-current-continuation} (a.k.a. \rackett{call/cc}) operator in ~\citet{ianjohnson:Vardoulakis2011Pushdown}.
%
This impressive work illuminated the fact that we can harness the enhanced technology of pushdown analyses in non-pushdown models of computation.
%
Doing this sacrifices call/return matching in the general case, but in practice the precision is much better than the alternative regular model that, say, AAM would provide.
%%

%%
A downside of the work providing \rackett{call/cc} is that it was an algorithmic change to the already complicated CFA2 --- there was no recipe for how to do this for one's favorite control operator.
%
This paper seeks to do just that with an operational view of what summarization is, in essence.
%
In other words, we give a concrete semantics to the tricks that the analyses in the literature use in the abstract, and maintain the original meaning of the language.
%
We also give intuitive analogies to well-established ideas/techniques so that the working semanticist can write a pushdown analysis for her language.
%
In order to demonstrate the applicability of this viewpoint, we show a new analysis for a language with composable control.
%
All of the semantics modeled in this paper are implemented in full detail in PLT Redex~\citep{ianjohnson:Felleisen:2009:SEP:1795772} and available online\footnote{\url{http://github.com/ianj/concrete-summaries}}.

@@

PDCFA does not have some orthogonal semantic components that CFA2 features to improve precision and is thus the simpler of the two.
%
The key to their method is to notice in languages without stack-capturing features, the continuation is only modified a frame at a time.
%
If the state space without the stack is finite (and it can be made finite with an approximation ala AAM), then the model falls directly into the realm of pushdown systems.
%
They thus recast the problem in terms of pushdown systems.
%
The whole of their machinery is then computing the pushdown system on-the-fly, only considering states reachable from the root (initial) state.
%
They call the main data-structure for this a Dyck state graph.
%
To match nodes that push frames with later nodes following the pop of that frame, they additionally compute an $\epsilon$-closure graph, which keeps edges between nodes that have a net-zero stack change path between them.
%
We skip the detour to pushdown systems altogether and show an operational understanding of the resulting analysis.
%%

%%
We start by deriving PDCFA from an operational semantics for the untyped lambda calculus (figure \ref{fig:base-semantics}).
%
The semantic spaces for the machine follow.

\begin{minipage}[b]{.55\linewidth}
  \begin{align*}
  \mexpr \in \Expr &= \svar{\mvar} \mid \sapp{\mexpr}{\mexpr} \mid \slam{\mvar}{\mexpr} \\
  \mstate \in \State &= (\Expr \times \Env) \times \Store \times \Kont \\
  \mval \in \Value &::= \vclo{\slam{\mvar}{\mexpr}}{\menv} \\
  \mkont \in \Kont &::= \kmt \mid \kcons{\mkframe}{\mkont} \\
  \mkframe \in \Frame &::= \kar{\mexpr,\menv} \mid \kfn{\mval}
  \end{align*}\hrule height 0pt\end{minipage}
\begin{minipage}[b]{.40\linewidth}
  \begin{align*}
  \menv \in \Env &= \Var \to \Addr \\
  \mstore \in \Store &= \Addr \to \wp(\Value) \\
  \mvar \in \Var &\text{ an infinite set} \\
  \maddr \in \Addr &\text{ an infinite set}
  \end{align*}\hrule height 0pt\end{minipage}

The $\Addr$ space is what controls the precision of the model.
%
For a concrete semantics, we require the allocation meta-function $\alloc :
\State \to \Addr$ to return fresh addresses ($\maddr \notin \dom(\mstore)$), but any choice of address is sound.
%
If $\alloc$ only uses addresses from a finite subset of $\Addr$, then the state space without continuations is finite, and the space of contination frames is finite (thus the pushdown system interpretation is apt).
%
This meta-function along with the commitment to having no recursive data-structures is the key to the AAM technique.
%
All recursion can be expressed by indirecting through addresses into a store --- tying Landin's knot.
%
We borrow this technique on top of our own.

\begin{figure}
  \centering
  $\mstate \stepto \mstate' \text{ where } \maddr = \alloc(\mstate)$ \\
  \begin{tabular}{r|l}
    \hline
% variable lookup
    $\tpl{(\svar\mvar, \menv), \mstore, \mkont}$
    &
    $\tpl{\mval,\mstore,\mkont}$ if $\mval \in \mstore(\menv(\mvar))$
    \\
% application
    $\tpl{(\sapp{\mexpri0}{\mexpri1}, \menv), \mstore, \mkont}$
    &
    $\tpl{(\mexpri0, \menv), \mstore, \kcons{\kar{\mexpri1,\menv}}{\mkont}}$
    \\
% argument evaluation
    $\tpl{\mval, \mstore, \kcons{\kar{\mexpr,\menv}}{\mkont}}$
    &
    $\tpl{(\mexpr, \menv), \mstore, \kcons{\kfn{\mval}}{\mkont}}$
    \\
% function call
    $\tpl{\mval,\mstore,\kcons{\kfn{\vclo{\slam{\mvar}{\mexpr}}{\menv}}}{\mkont}}$
    &
    $\tpl{(\mexpr, \extm{\menv}{\mvar}{\maddr}), \joinone{\mstore}{\maddr}{\mval}, \mkont}$
  \end{tabular}
  \caption{The CESK machine}
  \label{fig:base-semantics}
\end{figure}

\paragraph{Tracking return points:} the magic of the method is in keeping a table of continuation segments for each function and store pair, where the continuation is segmented at function boundaries.
%
This closely mirrors the technique of store-allocating continuations used in AAM.
%
Instead of deferring to the table (store) for the remainder of a continuation for each frame, we only have indirections at function call boundaries (see rules 4 and 5 in figure \ref{fig:summary-semantics} for saving and restoring continuations).
%
Note that since AAM is about finitizing the state space, this step itself fits within the AAM method, since continuations within functions truncated at call boundaries are bounded by the nesting depth of those functions, thereby making the continuation space finite.
%
In the monovariant case, the number of continuations is still linear in the size of the program.
%%

%%
The tail of a continuation in the context of a function will contain the stack-less context in which the function was called, in order to link up with the proper call-site(s).
%
The context includes the function (or unique label of the function), the environment, and the store.
%
Notice that this choice can be viewed as a special allocation strategy for continuations in the AAM viewpoint, but we separate the table of continuations from the store since continuations themselves contain stores in $\rt$ frames --- this leads to a recursive data-structure that we are trying to avoid.
%
The store is an important ingredient to maintaining enough precision to keep a pushdown abstraction.
%
The semantic spaces for the machine are modified thusly:

\begin{align*}
  \mstate \in \State &= (\Expr \times \Env) \times \Store \times \Kont \times \KTab \times \Memo \\
  \mkont \in \Kont &::= \kmt \mid \krt{(\mexpr,\menv),\mstore} \mid \kcons{\mkframe}{\mkont} \\
  \mktab \in \KTab &= (\Expr \times \Env)\times \Store \to \wp(\Kont) \\
  \mmemo \in \Memo &= (\Expr \times \Env)\times \Store \to \wp(\Value \times \Store)
\end{align*}

\paragraph{The role of summaries:} notice the additional $\Memo$ component.
%
A ``summary edge'' in CFA2 or equivalently an $\epsilon$-edge in PDCFA is an edge from the source of a push edge to the target of the matching pop edge.
%
``Matching'' here means there is a path through machine reductions that don't change the stack, or through summary edges.
%
In our model, we only ``push'' when we call a function, and we only ``pop'' when we return with a value.
%
There is an analogy to something more operational: summary edges embody memoization (see the last rule in figure \ref{fig:summary-semantics}).
%
Instead of following an entire path through a call to return a value, we simply jump from the call to the return with the result of the call, with the resulting context changes---in this case only the store (the second case of rule 4).

\begin{figure}
  \centering
  $\mstate \stepto \mstate' \text{ where } \maddr = \alloc(\mstate)$ \\
  \begin{tabular}{r|l}
    \hline
% variable lookup
    $\tpl{(\svar\mvar, \menv), \mstore, \mkont, \mktab, \mmemo}$
    &
    $\tpl{\mval,\mstore,\mkont, \mktab, \mmemo}$ if $\mval \in \mstore(\menv(\mvar))$
    \\
% application
    $\tpl{(\sapp{\mexpri0}{\mexpri1}, \menv), \mstore, \mkont, \mktab, \mmemo}$
    &
    $\tpl{(\mexpri0, \menv), \mstore, \kcons{\kar{\mexpri1,\menv}}{\mkont}, \mktab, \mmemo}$
    \\
% argument evaluation
    $\tpl{\mval, \mstore, \kcons{\kar{\mexpr,\menv}}{\mkont}, \mktab, \mmemo}$
    &
    $\tpl{(\mexpr, \menv), \mstore, \kcons{\kfn{\mval}}{\mkont}, \mktab, \mmemo}$
    \\
% function call
    $\tpl{\mval,\mstore,\kcons{\kfn{\vclo{\slam{\mvar}{\mexpr}}{\menv}}}{\mkont},\mktab,\mmemo}$
    & % actually do call,
    $\tpl{\mpoint,
          \mstore',
          \krt{\mpoint, \mstore'},
          \mktab',
          \mmemo}$
\\
    & or \\
    & $\tpl{\mval_\mathit{result},
            \mstore'',
            \mkont,
            \mktab',
            \mmemo}$
    \\ & \quad if $(\mval_\mathit{result},\mstore'') \in \mmemo(\mpoint,\mstore')$
    \\ % or, lookup in table
%    & $\tpl{\mval, \mstore', \mkont,\mktab,\mmemo}$ if $\mval \in \mmemo(\mpoint, \mstore')$ \\    
    where & $\mpoint = (\mexpr, \extm{\menv}{\mvar}{\maddr})$ \\
          & $\mstore' = \joinone{\mstore}{\maddr}{\mval}$ \\
          & $\mktab' = \joinone{\mktab}{(\mpoint, \mstore')}{\mkont}$
    \\
% return
    $\tpl{\mval, \mstore, \krt{\mpoint,\mstore'}, \mktab, \mmemo}$
    &
    $\tpl{\mval, \mstore, \mkont, \mktab, \joinone{\mmemo}{(\mpoint, \mstore')}{(\mval,\mstore)}}$
    \\ & \quad if $\mkont \in \mktab(\mpoint, \mstore')$
  \end{tabular}
  \caption{The summarizing tabular stack machine}
  \label{fig:summary-semantics}
\end{figure}

To turn this semantics into PDCFA's algorithm for constructing a Dyck state graph, given the appropriate $\alloc$, we apply a widening operator to make $\mstore$, $\mkont$, and $\mmemo$ shared amongst all states (meaning states then become represented without these compontents), in what is then called a $\System$.
%
Since these components are shared, they are always the least upper bound of any respective component that the semantics produces (the intermediate set $I$) in order to stay sound.
%
We use a meta-function $\wn$, ``wide to narrow'', to shift between the two representations of states.

%% FIXME: remove stores from memo-table since they will always be unchanged
%% TODO: Show that store widening and store/memo/KTab widening have a bisimulation.
\begin{align*}
  \mastate \in \sa{State} &= (\Expr \times \Env) \times \Kont \\
  \System &= \wp(\sa{State} \times \Store) \times \wp(\sa{State}) \times \Store \times \KTab \times \Memo \\
  {\mathcal F} &: \System \to \System \\
  {\mathcal F}(S, F, \mstore, \mktab, \mmemo) &= (S \cup S', F', \mstore', \mktab', \mmemo') \\
  \text{where } I &= \setbuild{\mstate'}{\mastate \in F, \wn(\mastate,\mstore,\mktab,\mmemo) \stepto \mstate'} \\
                \mstore' &= \bigsqcup\setbuild{\mstore'}{\wn(\_, \mstore',\_,\_) \in I} \\
                \mktab' &=  \bigsqcup\setbuild{\mktab'}{\wn(\_, \_, \mktab', \_) \in I} \\
                \mmemo' &=  \bigsqcup\setbuild{\mmemo'}{\wn(\_, \_, \_, \mmemo') \in I} \\
                S' &= \setbuild{(\mastate',\mstore')}{\wn(\mastate', \_, \_, \_) \in I} \\
                F' &= \setbuild{\mastate'}{(\mastate',\_) \in S' \setminus S} \\
                \wn(\tpl{(\mexpr,\menv),\mkont},\mstore,\mktab,\mmemo)
                   &= \tpl{(\mexpr,\menv),\mstore,\mkont,\mktab,\mmemo}
\end{align*}

A $\System$ embodies the states seen and at which store, $S$, the frontier set of states yet to analyze, $F$, the shared store for the frontier, $\mstore$, the continuation table $\mktab$ and the memo table $\mmemo$.
%
The memo table need only map to values, since the global store will always subsume the store that would otherwise be mapped.
%
All the states in the frontier are stepped with the current store, after which the next store is the least upper bound of all the resulting stores.
%
The next frontier contains only states resulting from stepping the previous frontier that we haven't seen at this next store.
%
Systematic techniques for a performant implementation can be found in \citet{ianjohnson:oaam:icfp2013}.

\paragraph{Example:} Let's consider a monovariant allocation strategy (the address for each binding is the variable name itself) to run the following example:
\begin{lstlisting}[mathescape]
  (let* ([id ($\lambda$ (x) x)]
         [y (id 0)]
         [z (id 1)])
    ($\le$ y z))
\end{lstlisting}
Suppose we extend our semantics to allow numbers, numeric primitives and \texttt{let}.
%
The continuation frame for a \texttt{let} contains the identifier to bind to the resulting value, along with the body of the \texttt{let} with its environment.
%
Call the constructor of this frame $\mathbf{lt}$.
%%

%%
We should expect that a pushdown analysis would predict this evaluates to true, and there are no loops in the program.
%
0CFA ~\citep{ianjohnson:Shivers:1991:CFA} claims there is a loop from the second call of \texttt{id} to the first, and thusly predicts this program evaluates to true or false.
%
PDCFA claims there is no loop, and depending on the implementation, that the result is true or false (paper), or just true (implemented)
\footnote{This is because the paper steps every seen state with the current store every iteration, but the implementation only steps states that need stepping.}.
%
Let's take a look at the evaluation after binding \texttt{id} ($\mstore_0 = [\texttt{id} \mapsto \set{\vclo{\slam{\texttt{x}}{\texttt{x}}}{\bot}}]$):

@@

We maintained enough context to distinguish the return points of \texttt{id} to not rebind \texttt{y} to 1.
%
When determining control flow through the expression, we consult $\mmemo$ to continue past function calls, so there is no confusion about a back edge from the second call to \texttt{id}.
%

\begin{theorem}
  Tabular semantics has a bisimulation with the standard semantics with comparable allocation strategies.
\end{theorem}

This follows directly from the invariants we prove of $\mktab$ and $\mmemo$ in the tabular semantics.
%
Allocation strategies are comparable if they produce equal addresses regardless of the differences in the state representations.
%
More formally, $\alloc$ and $\alloc^*$ are comparable if this implication holds:
\begin{mathpar}
  \inferrule{\mkont \in \unroll{\mktab}{\mkont^*}}
            {\alloc(\tpl{\mpoint,\mstore,\mkont}) = \alloc^*(\tpl{\mpoint,\mstore,\mkont^*,\mktab,\mmemo})}
\end{mathpar}

The $\mathit{unroll}$ function interprets what are all the valid continuations that $\mktab$ encodes for a given continuation that contains $\krt{\mctx}$:
\begin{mathpar}
  \inferrule{ }{\kmt \in \unroll{\mktab}{\kmt}} \quad
  \inferrule{\mkont \in \unroll{\mktab}{\mkont'}}{\mkframe:\mkont \in \unroll{\mktab}{\mkframe:\mkont'}} \quad
  \inferrule{\mkont^\circ \in \mktab(\mctx) \\
             \mkont \in \unroll{\mktab}{\mkont^\circ}}
            {\mkont \in \unroll{\mktab}{\krt{\mctx}}}
\end{mathpar}

The astute reader may find this definition is dubious, since there is a possibility for unguarded corecursion here.
%
A function with ill-founded recursion can lead to a table such that $\krt{\mctx} \in \mktab(\mctx)$.
%
The abstractions that we impose on our data might also cause this situation, even though the concrete execution might always terminate.
%
Our interpretation of $\stepto$ is in a fixed-point computation, so circularities are not a problem for termination.

\begin{lemma}[Stack irrelevance]\label{lem:irrelevance}
  $\hastail(\mtrace,\mkont)$ implies $\replacetail(\mtrace,\mkont,\mkont')$ is a valid trace.
\end{lemma}
\begin{mathpar}
  \inferrule{ }{\hastail(\epsilon,\mkont)} \quad
  \inferrule{ }{\hastail(\tpl{\mpoint,\mstore,\append{\mkont'}{\mkont}},\mkont)} \quad
  \inferrule{\hastail(\mtrace\mstate,\mkont) \quad
             \mstate \stepto \mstate' \quad
             \hastail(\mstate',\mkont)}
            {\hastail(\mtrace\mstate\mstate',\mkont)}
\end{mathpar}

\begin{align*}
  \replacetail(\tpl{\mpoint,\mstore,\append{\mkont'}{\mkont}},\mkont,\mkont'') &= \tpl{\mpoint,\mstore,\append{\mkont'}{\mkont''}} \\
  \replacetail(\epsilon,\mkont,\mkont'') &= \epsilon \\
  \replacetail(\mtrace\mstate,\mkont,\mkont') &= \replacetail(\mtrace,\mkont,\mkont')\replacetail(\mstate,\mkont,\mkont') \\
\end{align*}
